---
title: "R Notebook"
output: html_notebook
---
```{r}
library('Seurat')
library(tidyverse)
```

```{r}
NML_1 <- Read10X(data.dir = "/Users/bhava/Downloads/GSE132771_RAW/NM1")
NML_2 <- Read10X(data.dir = "/Users/bhava/Downloads/GSE132771_RAW/NML2")
NML_3 <- Read10X(data.dir = "/Users/bhava/Downloads/GSE132771_RAW/NML3")
```

```{r}
NML_1 <- CreateSeuratObject(counts = NML_1, project = "NML_1", min.cells = 3, min.features = 200)
NML_2 <- CreateSeuratObject(counts = NML_2, project = "NML_2", min.cells = 3, min.features = 200)
NML_3 <- CreateSeuratObject(counts = NML_3, project = "NML_3", min.cells = 3, min.features = 200)
```

```{r}
NML_1
```

```{r}
colnames(NML_1)
rownames(NML_1)
```


```{r}
# for metadata
view(NML_1)
```

```{r}
# Merge Objects

MergedNML <- merge(NML_1, y = c(NML_2, NML_3),
                   add.cell.ids = ls()[1:3],
                   project = 'MergedNML')
```


```{r}
view(MergedNML)
```

#Pre-Processing
##1 QC
##2 Data Normalization
##3 Feature Selection (Finding Highly Varible Features)


### QC

Low Quality Cells have Low feature and Count_Rna. We need to filter that.

Low Quality Cells often have high percentage of mitochodrial genes

```{r}
range(MergedNML$nFeature_RNA)
range(MergedNML$nCount_RNA)
```

Adding Percentage of Mitochondrial genes
```{r}
MergedNML <- PercentageFeatureSet(MergedNML, pattern = "^MT-", col.name = "percent.mt")
view(MergedNML)
range(MergedNML$percent.mt)
```
```{r}
view(MergedNML)
```

```{r}
range(MergedNML$percent.mt)
```

```{r Visualizing the Distribution}
VlnPlot(MergedNML, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
MergedNML <- subset(MergedNML, subset = nFeature_RNA > 500 & nFeature_RNA < 4000 & nCount_RNA < 20000 & percent.mt < 10)
```

```{r}
MergedNML
```
```{r}
VlnPlot(MergedNML, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

### Data Normalization

```{r}
MergedNML <- NormalizeData(MergedNML, normalization.method = "LogNormalize", scale.factor = 10000 )
```

#### Feature Selection
```{r Find Highly Variable Features}

MergedNML <- FindVariableFeatures(MergedNML, selection.method = "vst", nfeatures = 2000)

MergedNML
```
We do this to find features which have high expression in some cells and no expression in some cells

#### Data Scaling

We Usually run scaling the most varianble features.
```{r}
MergedNML <- ScaleData(MergedNML) # 2000 Identified Variable Features

all.genes <- rownames(MergedNML)

MergedNML <- ScaleData(MergedNML, features = all.genes)

View(MergedNML@commands)

```

### Dimensionality Reduction

```{r}
MergedNML <-RunPCA(MergedNML)
```
```{r}
DimPlot(MergedNML, reduction = "pca", dims = c(1,2))
```
```{r}
DimPlot(MergedNML, reduction = "pca", dims = c(1,10))
```
```{r}
ElbowPlot(MergedNML)
```
We use Elbow Plot to determine how many PCs did it take to capture majority of the signal. In this case, We can see that the first 9 PCs explain majority of the signal

```{r}
ElbowPlot(MergedNML, ndims = 50, reduction = "pca")
```
### Clustering
```{r}
#Takes first 20 PCs as Input from previous PCA and find features with similar Expression Pattern
MergedNML <- FindNeighbors(MergedNML, dims = 1:20)

MergedNML <- FindClusters(MergedNML, resolution = 0.1)

```
First we use 0.1 resolution which gives us 11 clusters so increase the resolution to 0.3 to find more clusters

```{r}
MergedNML <- FindClusters(MergedNML, resolution = 0.3)
```
#### Run  non-linear Dimensional reducation (UMAP)
```{r}
MergedNML <- RunUMAP(MergedNML, dims = 1:20)

DimPlot(MergedNML, reduction = "umap", label = TRUE, repel = TRUE)
```

#### Finding Marker Genes
```{r}
MergedNML <- JoinLayers(MergedNML)

# Now find markers aka Differential expression analysis
MergedNML_markers <- FindAllMarkers(MergedNML, 
                                     only.pos = TRUE,
                                     min.pct = 0.25,
                                     logfc.threshold = 0.25)

# View results
View(MergedNML_markers)
```

#### Top 10 marked genes
```{r}

top10 <- MergedNML_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

View(top10)
```

#### Top 3
```{r}
MergedNML_markers %>%
  group_by(cluster) %>%
  slice_max(n = 3, order_by = avg_log2FC) %>%
  select(cluster, gene) %>%
  group_by(cluster) %>%
  summarize(top_genes = paste(gene, collapse = ", "))
```
#### Marker gene verification
```{r}
# Verify your top markers visually
FeaturePlot(MergedNML, features = c(
  "SFTPA2",   
  "CD3D",     
  "S100A8", 
  "ACKR1",  
  "KLRF1",  
  "LUM",    
  "APOE",   
  "ACTG2"   
))
```

#### Find fold change values
```{r}
MergedNML_markers %>%
  group_by(cluster) %>%
  slice_max(n = 3, order_by = avg_log2FC) %>%
  select(cluster, gene, avg_log2FC) %>%
  print(n = 54) 
```

#### Annotaation
```{r}
new.cluster.ids <- c(
  "Monocytes",                # 0
  "Alveolar Macrophages",     # 1
  "Alveolar Epithelial",      # 2
  "Vascular Endothelial",     # 3
  "Alveolar Type 2",          # 4
  "Fibroblasts",              # 5
  "Dendritic Cells",          # 6
  "T Cells",                  # 7
  "NK Cells",                 # 8
  "Smooth Muscle",            # 9
  "Mesothelial",              # 10
  "Macrophages",              # 11
  "Rare Epithelial",          # 12
  "B Cells",                  # 13
  "Mast Cells",               # 14
  "Proliferating",            # 15
  "Stromal",                  # 16
  "Ciliated/Lymphatic"        # 17
)

names(new.cluster.ids) <- levels(Idents(MergedNML))
MergedNML <- RenameIdents(MergedNML, new.cluster.ids)

# Save annotated object
MergedNML$cell_type <- Idents(MergedNML)

# Plot final annotated UMAP
DimPlot(MergedNML, reduction = "umap", label = TRUE, pt.size = 0.5, label.size = 3) +
  ggtitle("Healthy Lung Cell Atlas - Annotated") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))


```

```{r}
total_cells <- ncol(MergedNML)
print(paste("Total cells analyzed:", total_cells))

```

